---
title: "DataViz"
author: "Kevin S.W."
date: "9/26/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message, FALSE)

Packages <- c("tidyverse", "ggridges", "hexbin")
invisible(lapply(Packages, library, character.only = TRUE))

```

# Data Viz 1
Creating ggplot from weather data from NOAA

## Loading weather data

```{r weather_data, cache = TRUE}

# Package interfaces with the weather station/service; interacts with NOAA servers and downloads the updated data
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"), # Stations to download from
                      var = c("PRCP", "TMIN", "TMAX"), # which variables i want
                      date_min = "2017-01-01",     #range of dates min - max
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df

```

## creating ggplot
```{r}

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

# alternative way to make the plot:
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

#saving initial plots
weather_df %>% filter(name == "CentralPark_NY") #without any variable assignments, this code will "print out" immediately.

# if we want to save instead, we can assign a variable to it. i.e.
scatter = weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

scatter
```

## Adding color to plot

```{r}

# Aesthetic maps can be added inside each "assignments" or just do it outside
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.4)
  

```

## why does aes() positions matter?

```{r}

# if we add geom_smooth at the end...
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.4) +
  geom_smooth(se = FALSE) # se = FALSE removes the "error bars" of the line plot


# if we move the aes up to ggplot...
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) +
  geom_smooth(se = FALSE) 

```

by adding aes() inside the geom_point(), it only "applies" to those variables. If we put it into ggplot however, it will apply to everything.

## facets!
manipulating the data presentation beyond 1 graph
```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) +
  geom_smooth(se = FALSE) +
  facet_grid(~name) # the thing on the right side of the tilde, will be columns

```

## lets do more plotting
```{r}

# we can see temp changes over time
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = tmax), alpha = 0.3) +  # adding size = prcp into point changes the sizes based on the "intensity".
  geom_smooth(se = FALSE)

# or if we want to see precip
weather_df %>% 
  ggplot(aes(x = date, y = prcp, color = name)) +
  geom_point() +  
  geom_smooth(se = FALSE)

```

## Some extra stuff

```{r}

weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = tmax), alpha = 0.3) +  
  geom_smooth(size = 2, se = FALSE)

```


## 2d density
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
  
```

## more plots

```{r}

## Examples of univariate plots (only x variable)

# color actually only colors the "outlines".
# so we need to use "fill ="
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_histogram(position = "dodge") #adding position = "dodge" allows the bars next to each other

# alternatively
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_histogram() +
  facet_grid(~name)

#density plot...
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_density(alpha = 0.3)

# boxplots
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) +  
  geom_boxplot()

# violin plots; usually you can see a "density" plot if you turn your head sideways
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) +  
  geom_violin()

# ridge plots; useful when you have "lots" of datapoints
weather_df %>% 
  ggplot(aes(x = tmax, y = name)) +  
  geom_density_ridges() # in ggridges

```
