---
title: "DataViz"
author: "Kevin S.W."
date: "9/26/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message, FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "90%")

Packages <- c("tidyverse", "ggridges", "hexbin", "patchwork")
invisible(lapply(Packages, library, character.only = TRUE))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

# Data Viz 1
Creating ggplot from weather data from NOAA

## Loading weather data

```{r weather_data, cache = TRUE}

# Package interfaces with the weather station/service; interacts with NOAA servers and downloads the updated data
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"), # Stations to download from
                      var = c("PRCP", "TMIN", "TMAX"), # which variables i want
                      date_min = "2017-01-01",     #range of dates min - max
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df

```

## creating ggplot
```{r}

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

# alternative way to make the plot:
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

#saving initial plots
weather_df %>% filter(name == "CentralPark_NY") #without any variable assignments, this code will "print out" immediately.

# if we want to save instead, we can assign a variable to it. i.e.
scatter = weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

scatter
```

## Adding color to plot

```{r}

# Aesthetic maps can be added inside each "assignments" or just do it outside
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.4)
  

```

## why does aes() positions matter?

```{r}

# if we add geom_smooth at the end...
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.4) +
  geom_smooth(se = FALSE) # se = FALSE removes the "error bars" of the line plot


# if we move the aes up to ggplot...
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) +
  geom_smooth(se = FALSE) 

```

by adding aes() inside the geom_point(), it only "applies" to those variables. If we put it into ggplot however, it will apply to everything.

## facets!
manipulating the data presentation beyond 1 graph
```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) +
  geom_smooth(se = FALSE) +
  facet_grid(~name) # the thing on the right side of the tilde, will be columns

```

## lets do more plotting
```{r}

# we can see temp changes over time
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = tmax), alpha = 0.3) +  # adding size = prcp into point changes the sizes based on the "intensity".
  geom_smooth(se = FALSE)

# or if we want to see precip
weather_df %>% 
  ggplot(aes(x = date, y = prcp, color = name)) +
  geom_point() +  
  geom_smooth(se = FALSE)

```

## Some extra stuff

```{r}

weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = tmax), alpha = 0.3) +  
  geom_smooth(size = 2, se = FALSE)

```


## 2d density
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
  
```

## more plots

```{r}

## Examples of univariate plots (only x variable)

# color actually only colors the "outlines".
# so we need to use "fill ="
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_histogram(position = "dodge") #adding position = "dodge" allows the bars next to each other

# alternatively
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_histogram() +
  facet_grid(~name)

#density plot...
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +  
  geom_density(alpha = 0.3)

# boxplots
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) +  
  geom_boxplot()

# violin plots; usually you can see a "density" plot if you turn your head sideways
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) +  
  geom_violin()

# ridge plots; useful when you have "lots" of datapoints
weather_df %>% 
  ggplot(aes(x = tmax, y = name)) +  
  geom_density_ridges() # in ggridges

```

# saving plots
```{r}

# to save a plot; assign a name to it
gg_ridge = 
  weather_df %>% 
  ggplot(aes(x = tmax, y = name)) +  
  geom_density_ridges() # in ggridges

# format is (filename; ie location of file, which plot). if location not specified, will be local folder
ggsave("ggplot_temp_ridge.pdf", gg_ridge) 

# possible to do ggsave() as it assumes the last plot...but this is dangerous practice.

```

# regarding boxplots and figure dimensions in RMarkdown

when specifying in the chunk the figure width (fig.width = ...) this sets the boundaries of the graph. The larger it is, the wider the boundaries.

```{r, fig.width = 6}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) + 
  geom_smooth(se = FALSE)

```

```{r, fig.width = 12}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(aes(alpha = 0.4)) +
  geom_smooth(se = FALSE)

```

# Data Viz II - ggplot more in depth

```{r weather_data2, cache = TRUE}

#caching saves the output into a certain code chunk...it's like "RAM". 

weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"), # Stations to download from
                      var = c("PRCP", "TMIN", "TMAX"), # which variables i want
                      date_min = "2017-01-01",     #range of dates min - max
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df

```

## making new plots

start with an old plot

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5)

```

then we add labels using labs()...

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  labs(
    title = "Temp plot",
    x = "min temp in C",
    y = "max temp in C",
    caption = "data from NOAA via rnoaa package"
  )

```

then we can adjust its scales. The form is scale_*  , such as x-axis tick marks, etc

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  labs(
    title = "Temp plot",
    x = "min temp in C",
    y = "max temp in C",
    caption = "data from NOAA via rnoaa package") +
  scale_x_continuous(
    breaks = c(-15,-5,20),   # x-axis "tick marks" (the guide ticks)
    labels = c("-15C", "-5", "-20")) +  # to change those specific "labels"
# transforms datapoint to be the "sqrt" of each points. alternatively you can have scale_y_sqrt()
   scale_y_continuous(
    trans = "sqrt")

```

## Colors!!

Setting your own hues...
```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  labs(
    title = "Temp plot",
    x = "min temp in C",
    y = "max temp in C",
    caption = "data from NOAA via rnoaa package") +
  scale_color_hue(         # this is setting the hues of the color yourself
    name = "Location",
    h = c(50, 200)           # range of the hue to take
  )
  
```

Or we can use `viridis` package.
This is super useful because the gradients in the package are made with color-blindness and grayscales in mind (they still retain the gradient).

```{r}

gg_base <- weather_df %>%    # making a baseline plot
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  labs(
    title = "Temp plot",
    x = "min temp in C",
    y = "max temp in C",
    caption = "data from NOAA via rnoaa package") +
  viridis::scale_color_viridis(
    name = "Location",
    discrete = TRUE
  )
```

more in depth about color...

```{r}

ggplot(aes(x = tmin, y = tmax, color = "red")) + 
  geom_point(alpha = .5) +

# adding "color =" in the ggplot...assigns that color to a specific variable.
# if we just wants to change the color, then we have to specify it in the geom_* variables.
  
```


## Themes...
themes() are "global" things that deal with the general things of the plot. Typically organizations

```{r}

gg_base +
  theme(
    legend.position = "bottom"    # replaces the position of the legend
  )

```

We can change the theme_*...such as theme_bw()

```{r}
gg_base +
  theme_bw()+
  theme(legend.position = "bottom")
```

or theme_minimal()...

NOTE: in ggplot, ORDER MATTERS!! i.e. applying general "themes" later will "reset" whatever granular modifications you did since it overrides prior commands.

Themes are recommended to be applied first before adjusting anything else. Check `ggthemes` for more themes.

```{r}

gg_base +
  theme_minimal() +
  theme(legend.position = "none")

```


you can add the code below at the top along with the initial code chunk in order to set it globally

```{r}

theme_set(theme_bw() + theme(legend.position = "bottom"))

```

if you have more than one dataset, you could combine data sets with different forms as long as you specify/make data frames out separately...for example:
```{r}

central_park = 
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki = 
  weather_df %>% 
  filter(name == "Waikiki_HA")

waikiki %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp)) +
  geom_line(data = central_park)

```

## Multi-panel plots...

Occasionally, you'd want to include different plots into the same "canvas". We can use patchwork...
```{r}

gg_scatter <-
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

gg_density <-
  weather_df %>% 
  ggplot(aes(x = tmin)) +
  geom_density()

gg_box <-
  weather_df %>% 
  ggplot(aes(x = name, y = tmax, color = name)) +
  geom_boxplot()

# + adds the plot to the side, / adds below. order of the operations matter
(gg_scatter + gg_density) / gg_box

```

## manipulating factor variables in ggplots...
When it comes to boxplot, the variables will be dependent on factors. the order of the factors are by default, based on alphabetical ordering.
```{r}

weather_df %>% 
  mutate(
    name = factor(name),
    name = fct_relevel(name, "Waikiki_HA", "CentralPark_NY")
  ) %>% 
  ggplot(aes(x = name, y = tmax, color = name)) +
  geom_boxplot()



```


reorder...
```{r}

weather_df %>% 
  mutate(
    name = factor(name),
    name = fct_reorder(name, tmax) # reorders according to the y variable
  ) %>% 
  ggplot(aes(x = name, y = tmax, color = name)) +
  geom_boxplot()

```

so in this case, since waterhole's tmax are the lowest, it automatically becomes the first "factor". 


## making multiple plots with the same variables and keeping consistency between...

first, we have to restructure then plot...

i.e. if we try to make 3 separate density plots with the same variables and separation of "factors"...
```{r}
weather_df %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation",
    values_to = "temperature"
  ) %>% 
  ggplot(aes(x = temperature, fill = observation)) +
  geom_density(alpha = 0.3) +
  facet_grid(~name) +  # splits the graph by name; left ~ right -> left = vert_split, right = hor_split
  theme(legend.position = "bottom")

```

another example with FAS_pups dataset...

```{r}

pup_data = 
  read_csv("./data/FAS_pups.csv", col_types = "ciiiii") %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `1` = "male", `2` = "female")) 

litter_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  mutate(wt_gain = gd18_weight - gd0_weight,
         day_of_tx = as.numeric(day_of_tx))

fas_data = left_join(pup_data, litter_data, by = "litter_number") 

fas_data %>% 
  select(sex, dose, day_of_tx, pd_ears:pd_walk) %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome", 
    values_to = "pn_day") %>% 
  drop_na() %>% 
  mutate(outcome = forcats::fct_reorder(outcome, day_of_tx, median)) %>% 
  ggplot(aes(x = dose, y = pn_day)) + 
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome)

```

# DataViz III/Exploratory Data Analysis

Moment where you mess with data and get a sense of your data. We can use group_by() or summarize(). Grouping doesn't change your data but it adds a layer on top of it. Summarize() gives 1-number summaries. You can also nest this, because if you start with group_by() and then summarize, the result is still a dataframe.

**Warning**: exploratory analyses are considered secondary analyses as these are usually done by "poking around" the data. This is typically not pre-planned and therefore shouldn't be prioritized unless the primary purpose has been completed. 


```{r loading_weather, cache = TRUE}

weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>% 
    # this is rounding down date to the beginning of its month
  select(name, id, date, month, everything())

```

## Group_by and counting things

```{r}

# it will look the same, but now if we check the weather_df, there is "Groups: name, month [total groups]
weather_df %>% 
  group_by(name, month)

# counting

weather_df %>% 
  group_by(month, name) %>% 
  summarize(n_obs = n()) # n() just counts; assigned to a new column and makes a new column

weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())

weather_df %>% 
  group_by(name) %>% 
  summarize(n_obs = n())


weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n(),
            n_unique = n_distinct(date)) # adding n_distinct "counts" the distinct dates

```

There are also other ways to count

```{r}

# instead of doing group_by then summarize, we can use count()
weather_df %>% 
  count(name)

weather_df %>% 
  count(name, month)  # by doing name = "something", it will also assign a name for that "new column"


# base R also has a counting function but this is horrible...
# because this isn't a dataframe
weather_df %>% 
  pull(name) %>% 
  table()

```

## digression...2x2 tables

```{r}

weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(            #case_when useful when you have new variables based on several prior variables 
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE ~ ""
    )  
  ) %>% # now we want to make 2x2 table
  group_by(name, cold) %>% 
  count() %>% 
  pivot_wider(
    names_from = "cold",
    values_from = "n"
  )


# but there is another way to get this as well...
# a note by jeff: if you need to find odds ratio...just use logistic regression
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(            #case_when useful when you have new variables based on several prior variables 
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE ~ ""
    )  
  ) %>%
  janitor::tabyl(name, cold)

```

## general summaries

```{r}

weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE),
    median_prcp = median(prcp, na.rm = TRUE),
  ) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() + geom_line()

# you'll notice that waikiki has NA...
weather_df %>% 
  filter(is.na(tmax)) # now that we saw some NA, we can add na.rm = TRUE in prior code...

```


## revisiting "2x2" table...
occasionally, your table won't look "pretty" for people and therefore may require pivot_wider...
```{r}

weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
    ) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  )

```

## Grouped mutates
**Warning**: this is a portion where grouping and forgetting can be troublesome...therefore there is a way to ungroup it...by using ungroup()

```{r}
# by having group_by prior to mutate, now mutate() applies to the assigned group.
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()

```

## Occasionally..you'll need window functions in grouped mutates...

```{r}

# we can get a ranked order of something...and filter only by certain ranks...

weather_df %>% 
  group_by(name, month) %>% 
  mutate(
    tmax_rank = min_rank(tmax)
  ) %>% 
  filter(tmax_rank == 1)

# lags and leads...

#e.g. how much does tmax variable change by day
# lagging/leading typically is used for predictive models based on "prior events".
weather_df %>% 
  group_by(name) %>%  # group by name b/c we want to be location specific
  mutate(
    lagged_tmax = lag(tmax),       # lag() is going to offset by one, 
    one_day_tmax_change = tmax - lagged_tmax
  ) %>% 
  summarize(sd_daily_change = sd(one_day_tmax_change, na.rm = TRUE))

```

